<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{The Visual Raster Cheat Sheet}
-->

The Visual Raster Cheat Sheet
========================================================

#### Etienne B. Racine (etiennebr@gmail.com)
#### 14 may 2015

***

```{r echo = FALSE, cache = FALSE}
library("knitr")
library("xtable")
opts_chunk$set(dev = c("png"), 
               fig.height = 4, 
               fig.width = 4, 
               warning = FALSE
               )
# modify stylesheet
 options(markdown.HTML.stylesheet = system.file('misc', 'vr_vignette.css', package='visualraster')) 
# xtable params
options(xtable.type = "html")
options(xtable.include.rownames = FALSE)
```


```{r echo = FALSE, cache = FALSE, message = FALSE}
library("visualraster")
library("igraph")    # boundaries and clumps
library("fields")    # Tps for interpolation, predict.se
library("maps")      # world data
library("mapproj")   # projections in ggplot2
library("rgeos")
library("raster")

library("reshape")

library("grid")
library("ggplot2")
library("RColorBrewer")

# define palette of colors
pal = list (
  qualitative = brewer.pal(11, "Set3"),
  circular = c("#74A9CF", "#BDC9E1", "#FFFFFF", 
               "#FEF0D9", "#FDCC8A", "#E34A33",
               "#B30000", "#000000", "#045A8D", 
               "#2B8CBE"),
  circular2 = c("#0088FF", "#ffffff", "#F51D1D", "#400038", "#0088FF")
)
names(pal$qualitative) = 1:length(pal$qualitative)

proj_random_tm = "+proj=tmerc +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +units=m"

# set fullframe as default theme
theme_set(theme_fullframe())

# make randomness repeatable
set.seed(101)
```


# I/O
Input and output operations, load or write data

## raster
```{r}
r = raster(matrix(sample(1:9, 100, replace = TRUE), 10, 10))
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r) + vr_geom_text(r)
```

## write
```{r eval = FALSE}
writeRaster(r, "raster.tif")
```

## read
```{r eval = FALSE, echo = FALSE}
# this chunk if for house keeping when running chunks without knitr
.r = r
```

```{r eval = FALSE}
r = raster("raster.tif")
```


```{r eval = FALSE, echo = FALSE}
# this chunk if for house keeping when running chunks without knitr
file.remove("raster.tif")
r = .r
```

# Local
Local analysis consider cells independently

## Access
Direct index `r[i]`

```{r echo = FALSE}
# generate point
pt = data.frame(x=0.22, y=0.72)
gpt = vr_geom_point(pt)

# generate line
stream = data.frame(
  x = cumsum(runif(7, max = 0.2)),
  y = cumsum(runif(7, max = 0.2))
)

ln = SpatialLines(list(Lines(list(Line(stream)), "1")) )
gln = vr_geom_line(stream)

# generate polygon
poly = SpatialPolygons(list(Polygons(list(Polygon(
  rbind(c(0.32, 0.74), 
        c(0.62, 0.84), 
        c(0.72, 0.25), 
        c(0.44, 0.25), 
        c(0.32, 0.74)))), 1)))
cpp  = fortify(poly)
gpy = vr_geom_polygon(cpp)
```

```{r echo = FALSE}
ggplot() + 
  vr_geom_raster(r) + 
  vr_geom_text(r, label = 1:100, family = "mono", size = 3)
```

Using line and columns `r[lin, col]`

```{r echo = FALSE}
grd = expand.grid(col = 1:10, lin = 1:10)
ref = with(grd, paste0(lin, ",", col))
ggplot() + 
  vr_geom_raster(r) +
  vr_geom_text(r, family = "mono", size = 3, label = ref)
```

```{r echo = FALSE}
# plot geometries
ggplot() + vr_geom_raster(r) + 
  vr_geom_text(r, family = "mono", size = 3, label = 1:100) +
  gln + gpt + gpy
  
```

### Cell from point
```{r}
cellFromXY(r, pt)
colFromX(r, pt$x)
rowFromY(r, pt$y)
fourCellsFromXY(r, as.matrix(pt))
```

### Cell from line
```{r}
cellFromLine(r, ln)
```

### Cell from polygon
```{r}
cellFromPolygon(r, poly)
```

## aggregate
```{r}
r_agg = aggregate(r, 2)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_agg) + vr_geom_text(r_agg)
```
## disaggregate
```{r}
r_dis = disaggregate(r, 2)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(round(r_dis)) + vr_geom_text(r_dis)
```

## cover
```{r}
r_missing = raster(matrix(NA, 10, 10))
r_missing[41:60] = 6
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_missing) + vr_geom_text(r_missing)
```

```{r}
r_covered = cover(r_missing, r)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_covered) + vr_geom_text(r_covered)
```

## mask
```{r}
r_masked = mask(r, r_missing)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_masked) + vr_geom_text(r_masked)
```

## calc
```{r}
fun = function(x) { x * 10 }
r_mul = calc(r, fun)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_mul / 10) + vr_geom_text(r_mul)
```

## overlay
```{r}
r1 = init(r, function(x) sample(5:6, x, replace = TRUE))
r2 = init(r, function(x) sample(1:2, x, replace = TRUE))
r3 = overlay(r1, r2, fun=function(x,y){return(x+y)})
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r1) + vr_geom_text(r1)
ggplot() + vr_geom_raster(r2) + vr_geom_text(r2)
ggplot() + vr_geom_raster(r3) + vr_geom_text(r3)
```

## distance
We need a raster containing origins to compute distances (so we need units)
```{r}
r_origin = raster(matrix(NA, 10, 10))
extent(r_origin) = extent(c(0, 10, 0, 10))
projection(r_origin) = "+proj=tmerc +lat_0=0 +lon_0=0 +x_0=0 +y_0=0 +units=m"
r_origin[c(15, 16, 45)] = 1
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_origin) + vr_geom_text(r_origin)
```

```{r}
r_dist = distance(r_origin) 
```

```{r echo = FALSE}
ggplot() + vr_geom_raster_seq(r_dist) + vr_geom_text(r_dist)
```

## reclassify
```{r}
m = cbind(
  from    = c(0, 2, 4), 
  to      = c(2, 4, 6), 
  becomes = c(1, 2, 3)
  )
```

```{r echo = FALSE, results = 'asis'}
xtable(m, digits = 0)
```

```{r}
r_reclass = reclassify(r_dist, m)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_reclass) + vr_geom_text(r_reclass)
```

# Focal
Focal analysis consider a cell plus its direct neighbours in a contiguous and symetrical manner

## focal
```{r}
r_focal = focal(r, fun = mean, w = matrix(1/9,nrow=3,ncol=3))
```

```{r, echo = FALSE}
ggplot() + vr_geom_raster_seq(r) + vr_geom_text(r)
ggplot() + vr_geom_raster_seq(r_focal) + vr_geom_text(r_focal)
```

You can remove edge effect by ignoring NA values
```{r}
r_focal_nedge = focal(r, fun = mean, w = matrix(1/9,nrow=3,ncol=3), na.rm = TRUE, pad = TRUE)
```

```{r, echo = FALSE}
ggplot() + vr_geom_raster_seq(r_focal_nedge) + vr_geom_text(r_focal_nedge)
```

# Zonal
Zonal analysis consider group of cells in an irregular, but conitguous (in space or value) manner.

## zonal
```{r echo = FALSE}
m = cbind(
  from    = c(0, 2, 4), 
  to      = c(2, 4, 8), 
  becomes = c(1, 2, 3)
  )
z = reclassify(r_dist, m)
extent(z) = extent(r)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(z) + vr_geom_text(z)
ggplot() + vr_geom_raster(r) + vr_geom_text(r)
```

```{r eval = FALSE}
zonal(r, z, mean)
```

```{r echo = FALSE, results='asis'}
zones = zonal(r, z, mean)
colnames(zones) = gsub("structure.*$", "mean", colnames(zones))
xtable(zones, digits = c(0, 0, 1))
```

## clump
```{r echo = FALSE}
r_bin = raster(matrix(round(runif(100, max = 0.69)), 10, 10))
```

```{r}
r_clumps = clump(r_bin)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_bin) + vr_geom_text(r_bin)
ggplot() + vr_geom_raster(r_clumps) + vr_geom_text(r_clumps)
```

## boundaries
```{r}
require("igraph")
r_clus = init(r, function(x) ifelse(runif(x) > 0.2, 1, NA))
r_boundaries = raster::boundaries(r_clus)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_clus) + vr_geom_text(r_clus)
ggplot() + vr_geom_raster(r_boundaries) + vr_geom_text(r_boundaries)
```

## extract
```{r echo = FALSE}
# plot geometries, this time with values
ggplot() + vr_geom_raster(r) + vr_geom_text(r) +
  gpt +
  gln +
  gpy
```

```{r}
extract(r, pt)
extract(r, ln)
extract(r, poly)
```

## rasterize
```{r, eval = FALSE}
rasterize(pt, r)
rasterize(ln, r)
rasterize(poly, r)
```

```{r echo = FALSE}
r_pt   = rasterize(pt, r)
r_line = rasterize(ln, r)
r_poly = rasterize(poly, r)

ggplot() + vr_geom_raster(r_pt) + gpt
ggplot() + vr_geom_raster(r_line) + gln
ggplot() + vr_geom_raster(r_poly) + gpy
```


# Statistical
Statistical operations summarize the raster information

## density
```{r, eval = FALSE}
density(r)
```

```{r echo = FALSE}
qplot(r[], geom = "density", fill = I("grey20")) + theme_fullframe()
```

## histogram
```{r, eval = FALSE}
hist(r)
```

```{r echo = FALSE}
qplot(r[], geom = "histogram", fill = I("grey20"), binwidth = 1) + 
  theme_fullframe()
```

## Spatial autocorrelation
```{r eval = FALSE}
Moran(r)
Geary(r)
```

```{r, echo = FALSE}
label_me = function (x, dec = 2) {
  paste0(
    "Moran = ", round(Moran(x), dec), "\n", 
    "Geary = ", round(Geary(x), dec)
    )
}
autocor_plot = function(x) {
  ggplot() +
  list(vr_geom_raster_seq(x),
       vr_geom_text(x, size = 2),
       ggtitle(label_me(x)))
}

r_plain = init(r, function(x) 1)
r_plain[c(55, 46)] = c(4, 0)

r_split = r
r_split[1:50] = 1
r_split[51:100] = 2

the_rs = list(r, 
       r_dist, 
       r_plain, 
       r_split, 
       init(r, fun = function(x) rep(1:2, length = x)),
       init(r, fun = function(x) rep(1:3, length = x)),
       init(r, fun = function(x) rep(1:4, length = x)),
       init(r, fun = function(x) rep(1:5, length = x)),
       init(r, fun = function(x) rep(1:6, length = x))
       )

for (w in the_rs) print(autocor_plot(w))
```

# Geometric
## extent
```{r eval = FALSE}
extent(r)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r) +
geom_polygon(data = extent_as_poly(r), aes(x = x, y = y), fill = "#ffffffaa", size = 2, colour = "black", linetype = "dashed")
```

## crop
```{r eval = TRUE}
f = extent(r) - c(0.35, 0.43)
```

```{r eval = FALSE}
crop(r, f)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r) +
geom_polygon(data = extent_as_poly(f), aes(x = x, y = y), fill = "#ffffffaa", size = 2, colour = "black", linetype = "dashed")

r_crop = crop(r, f)

ggplot() + vr_geom_raster(crop(r, f)) + 
  geom_polygon(data = extent_as_poly(f), 
               aes(x = x, y = y), 
               fill = NA, size = 2, 
               colour = "black",
               linetype = "dashed") + 
  geom_polygon(data = extent_as_poly(r), 
               aes(x = x, y = y), fill = NA, size = NA)
```

## intersect
```{r}
g = extent(0.43, 1.1, -0.1, 0.53)
```

```{r eval = FALSE}
intersect(r, g)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r) + vr_geom_text(r, label = 1:100, family = "mono", size = 3) +
  geom_polygon(data = extent_as_poly(g), 
               aes(x = x, y = y), 
               fill = NA, 
               size = 2, colour = "#111111", linetype = "dashed")

ggplot() + 
  geom_polygon(data = extent_as_poly(extent(r)), aes(x = x, y = y), fill = NA, size = 2, colour = "#aaaaaaaa", linetype = "dashed") + theme_fullframe() + coord_equal() +
  vr_geom_raster(intersect(r, g)) + 
  geom_polygon(data = extent_as_poly(g), aes(x = x, y = y), fill = NA, size = 2, colour = "black", linetype = "dashed")
  
```

### cellsFromExtent
```{r}
cellsFromExtent(r, g)
```


## union
```{r eval = FALSE}
union(extent(r), g)
```

```{r echo = FALSE}
r_union = union(extent(r), g)

ggplot() + 
  geom_polygon(data = extent_as_poly(g), aes(x = x, y = y), fill = NA, size = 2, colour = "#aaaaaaaa", linetype = "dashed") +
  geom_polygon(data = extent_as_poly(r), aes(x = x, y = y), fill = NA, size = 2, colour = "#aaaaaaaa", linetype = "dashed") +
geom_polygon(data = extent_as_poly(r_union), aes(x = x, y = y), fill = NA, size = 2, colour = "#111111", linetype = "dashed")  + coord_equal()
```


## flip
```{r eval = FALSE}
flip(r_split, "y")
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_split)

ggplot() + vr_geom_raster(flip(r_split, "y"))
```

## extend
```{r eval = FALSE}
extend(r, 3)
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(extend(r, 3))
```

## mosaic

```{r echo = FALSE}
# prepare data
r1 = shift(r, x = 0.3, y = 0.2)
r1[] = 1
r2 = shift(r, x = -0.6, y = -0.6)
r2[] = 1
```

```{r eval = FALSE}
mosaic(r, r1, r2, fun = sum)
```

```{r echo = FALSE}
r_mos = mosaic(r, r1, r2, fun = sum, na.rm = TRUE)
r_mos[is.nan(r_mos)] = NA
# plot rasters seperately
mos_grid = geom_tile(data = as_data(r_mos), 
            aes(x = x, y = y, fill = NA), color = "#aaaaaa")

ggplot() + vr_geom_raster(r) + mos_grid
ggplot() + vr_geom_raster(r1) + mos_grid
ggplot() + vr_geom_raster(r2) + mos_grid
```

```{r echo = FALSE}
ggplot() + vr_geom_raster(r_mos) + vr_geom_text(r_mos, size = 2)
```

## projection
```{r eval = FALSE}
projection(r)
```

```{r echo = FALSE}
r_degrees = raster(nrow=9, ncol=18)
r_degrees[] = as.vector(t(matrix(1:ncell(r_degrees), nrow=nrow(r_degrees))))
```

```{r echo = FALSE}
require("maps")
world = map_data("world")

gw = list(
  geom_polygon(data=world, aes(x=long, y=lat, group=group), 
               colour="#111111", fill = "#ffffff22", size = 0.2),
  scale_y_continuous(breaks = NULL),
  scale_x_continuous(breaks = c(-180, 0, 180))
  )
```

```{r echo = FALSE}
require("mapproj")
prj = coord_map(projection="albers", lat0=30, lat1=60, xlim = c(-1, 1) * 170)
# to make the plot clean, you need to remove at least on row of pixels
crop_pixels = function(r, x) {
  xdim = (extent(r)@xmax- extent(r)@xmin) / nrow(r)
  ydim = (extent(r)@ymax- extent(r)@ymin) / ncol(r)
  crop(r, extent(r) - c(xdim, ydim) * x)
}
ggplot() + vr_geom_raster_seq(crop_pixels(r_degrees, c(1, 0))) + gw + prj
```

## rotate
```{r eval = FALSE}
rotate(r)
```

```{r echo = FALSE}
extent(r_degrees) = extent(0, 360, -90, 90)
ggplot() + vr_geom_raster_seq(r_degrees) + gw + theme_geographic() 
ggplot() + vr_geom_raster_seq(rotate(r_degrees)) + gw + theme_geographic() 
```

## rasterToPoints
```{r}
pts = rasterToPoints(r)
```

```{r echo = FALSE, message = FALSE}
pts = as.data.frame(pts)
pts$coords = with(pts, paste0(round(x * 100), ",", round(y * 100)))
ggplot() +
  vr_geom_raster(r) +
  vr_geom_raster(r, color = "#ffffff", fill = "#ffffffaa") +
  scale_color_manual(values = pal$qualitative) +
  geom_point(data = pts, aes(x = x, y = y, fill = factor(layer), label = coords), size = 6, shape = 21, color = I("#050505")) + 
  geom_point(data = pts, aes(x = x, y = y, color = factor(layer), label = coords), alpha = 0.5, size = 6, color = I("#050505"), shape = "+") + 
  geom_text(data = pts, aes(x = x, y = y, color = factor(layer), label = coords), size = 2, hjust = 0, vjust = 0, color = "#111111", angle = 45) +
  coord_equal() +  theme_fullframe()
```

## rasterToPoints
```{r}
pols = rasterToPolygons(r_clumps, dissolve = TRUE)
```

```{r echo = FALSE, message = FALSE}
pols_fortify = fortify(pols)

ggplot() + vr_geom_raster(r_clumps, color = "#111111") +
  geom_polygon(data = pols_fortify, aes(x = long, y = lat, group = group, color = id), fill = NA, size = 2) +
  geom_polygon(data = pols_fortify, aes(x = long, y = lat, group = group), fill = NA, size = 2, color = I("#00000055")) +
  scale_color_manual(values = pal$qualitative) 
```

# Terrain
```{r echo = FALSE}
r_volcano = raster(volcano)
extent(r_volcano) = extent(0, ncol(r_volcano),0, nrow(r_volcano))
projection(r_volcano) =  proj_random_tm
```

```{r echo = FALSE}
ggplot() + vr_geom_raster_seq(r_volcano)
```

## terrain
```{r}
r_terrain = terrain(r_volcano, opt = c("slope", "aspect", "tpi", "tri", "roughness", "flowdir"))
```


```{r echo = FALSE}
for(i in names(r_terrain)) 
  print(ggplot() + vr_geom_raster_seq(r_terrain[[i]]) + ggtitle(i)) 
```

Angular data can sometimes be better expressed using a circular palette. In the following figure, blue is an North orientation, while South is red. Both colors reach black at West and white at East. The preceeding figure had some sharp edges
on Noth faces, when angle slightly changed from 360 to 0.

```{r echo = FALSE, message = FALSE}
ggplot() + vr_geom_raster_seq(trim(r_terrain[["aspect"]])) + ggtitle("aspect") + scale_fill_gradientn(colours = pal$circular2, limits = c(0, 2 * pi)) 

ggplot() + vr_geom_raster_seq(power_to_angle(trim(r_terrain[["flowdir"]]))) + ggtitle("flowdir") + scale_fill_gradientn(colours = pal$circular2, limits = c(0, 360)) 
```


## hilshade
```{r}
r_hill = hillShade(r_terrain[["slope"]], r_terrain[["aspect"]])
```

```{r echo = FALSE}
ggplot() + vr_geom_raster_seq(r_hill) + ggtitle("Hillshade") 
```

## interpolate
```{r, echo = FALSE}
require("fields")
survey = as.data.frame(r_volcano, xy = TRUE)
survey = survey[sample(1:nrow(survey), 50), ]
```

```{r, eval = TRUE}
tps = Tps(survey[, c("x", "y")], survey$layer)
r_interpolate = interpolate(r_volcano, tps)
```

```{r echo = FALSE, message = FALSE}
vr_geom_raster_div = function (r, ...) {
  list(
    vr_geom_raster_seq(r, ...),
    scale_fill_gradientn(colours = c("#D53E4F", "#FFFFBF", "#3288BD"))
    )
  } 
gsurvey = vr_geom_point(survey, size = 2)
ggplot() + vr_geom_raster_seq(r_volcano) + ggtitle("Volcano") + gsurvey
ggplot() + vr_geom_raster_seq(r_interpolate, limits = range(r_volcano[])) + ggtitle("Interpolation") + gsurvey
ggplot() + vr_geom_raster_div(r_interpolate - r_volcano) + ggtitle("Differences") + gsurvey
ggplot() + vr_geom_raster_seq(interpolate(r_volcano, tps, fun=predictSE)) + ggtitle("Standard Error") + gsurvey
```

# layerize
```{r}
s = layerize(r)
```

```{r echo = FALSE}
dd = as_data(s, melt = TRUE)
levels(dd$variable) = gsub("X", "", levels(dd$variable))
dd$fill = factor(with(dd, as.integer(variable) * value))
dd$fill[dd$fill == 0] = NA
dd$fill = droplevels(dd$fill)
ggplot() +
  scale_fill_manual(values = pal$qualitative) +
  geom_tile(data = dd, 
            aes(x = x, y = y, fill = fill), 
            color = "#888888", size = 0.25) +
  facet_wrap(~variable) + coord_equal()
```
